---
title: "`r params$title`"
date: now
date-format: "YYYY-MM-DD HH:mm"
params:
  title:      "BOEM Marine Sensitivity Report"
  areas_json: "[]"
  ver:        "v6"
  format:     "html"
  api_base:   "https://api.marinesensitivity.org"
  mapsp_base: "https://shiny.marinesensitivity.org/mapsp"
format:
  html:
    toc: true
    smooth-scroll: true
    embed-resources: true
    format-links: false
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    lof: true
    lot: true
  docx:
    # reference-doc: "../workflows/libs/ESP Report Template 2025_1 tech.docx"
    toc: true
    toc-depth: 2
    lof: true
    lot: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(msens)
  library(sf)
  library(dplyr)
  library(terra)
  library(purrr)
  library(jsonlite)
  library(scales)
  library(stringr)
  library(yaml)
  library(knitr)
})

is_html    <- params$format == "html"
api_base   <- params$api_base
mapsp_base <- params$mapsp_base

areas <- jsonlite::fromJSON(params$areas_json, simplifyVector = FALSE)
if (length(areas) == 0) {
  stop("No areas provided. Pass `areas_json` as a JSON array of ",
       '[{"label":..,"kind":"wkt"|"pra","value":..}].')
}

# Program Area polygons: read the versioned gpkg matching params$ver —
# same source used to build the public pmtiles, but without tile
# fragmentation issues.
dir_data <- switch(
  Sys.info()[["sysname"]],
  "Darwin" = "~/My Drive/projects/msens/data",
  "Linux"  = "/share/data")
pra_gpkg <- file.path(
  dir_data, "derived", params$ver,
  paste0("ply_programareas_2026_", params$ver, ".gpkg"))
ply_pra <- if (any(purrr::map_chr(areas, "kind") == "pra")) {
  sf::st_read(pra_gpkg, quiet = TRUE)
}

area_to_sf <- function(a) {
  if (a$kind == "pra") {
    ply_pra |>
      dplyr::filter(programarea_key == a$value) |>
      sf::st_geometry() |>
      sf::st_as_sf()
  } else {
    sf::st_as_sfc(a$value, crs = 4326) |>
      sf::st_as_sf()
  }
}

con       <- msens::sdm_db_con(version = params$ver, read_only = TRUE)
# r_cell_id is only needed for drawn polygons (kind == "wkt") — skip
# the raster load if every area is a Program Area.
any_wkt   <- any(purrr::map_chr(areas, "kind") == "wkt")
r_cell_id <- if (any_wkt) msens::cell_id_raster()[["cell_id"]]

# Program Areas use the precomputed zone metrics / zone_cell lookups
# (msens::scores_for_pra, msens::cells_in_pra) so we skip the
# terra::rasterize() + cell aggregation path entirely. Drawn polygons
# still need the polygon-intersection pipeline.
areas_data <- purrr::map(areas, function(a) {
  ply <- area_to_sf(a)
  if (identical(a$kind, "pra")) {
    cells  <- msens::cells_in_pra(con, a$value)
    scores <- msens::scores_for_pra(con, a$value)
  } else {
    cells  <- msens::cells_in_polygon(ply, r_cell_id)
    scores <- msens::scores_for_cells(con, cells)
  }
  list(
    label  = a$label,
    kind   = a$kind,
    value  = a$value,
    ply    = ply,
    cells  = cells,
    scores = scores,
    spp    = msens::species_for_cells(con, cells))
})

# combined sf used by the Map section — carries label + mean score
# so the Map can color-code areas by score and label them in place.
all_areas <- do.call(
  rbind,
  purrr::map(areas_data, function(ad) {
    sf::st_sf(
      label    = ad$label,
      score    = round(msens::mean_score(ad$scores), 1),
      geometry = sf::st_geometry(ad$ply))
  }))

# color ramp matches the mapgl app's Spectral fill; range adapts to
# the areas in this report so a small score spread still uses the
# full ramp instead of collapsing to one band.
score_cols <- rev(RColorBrewer::brewer.pal(11, "Spectral"))
score_rng  <- range(all_areas$score, na.rm = TRUE)
if (diff(score_rng) == 0) score_rng <- score_rng + c(-0.5, 0.5)

# shared table emitter: responsive HTML or booktabs PDF/docx
emit_table <- function(df, caption, align = NULL) {
  if (is_html) {
    tbl <- knitr::kable(
      df,
      format     = "html",
      escape     = FALSE,
      caption    = caption,
      align      = align,
      table.attr = 'class="table table-sm"')
    cat('<div style="overflow-x: auto; margin-bottom: 1em;">\n\n')
    cat(as.character(tbl), sep = "\n")
    cat("\n\n</div>\n\n")
  } else {
    cat(knitr::kable(
      df,
      caption   = caption,
      align     = align,
      booktabs  = TRUE,
      longtable = TRUE), sep = "\n")
    cat("\n\n")
  }
}
```

This report was generated via input parameters submitted by the app at [app.marinesensitivity.org/mapgl](https://app.marinesensitivity.org/mapgl). For more information about this project please visit [MarineSensitivity.org](https://MarineSensitivity.org), where you can find the latest documentation, apps and log of changes.

::: {.callout-note collapse="true"}
## Parameters

```{r}
#| label: print-params
# show only the parameters that were actually submitted — derived
# summaries (n_cells, mean_score, component scores) now live in the
# Table of Scores below.
cat(yaml::as.yaml(list(
  title  = params$title,
  ver    = params$ver,
  format = params$format,
  areas  = purrr::map(areas_data, function(ad) {
    info <- list(label = ad$label, kind = ad$kind)
    if (identical(ad$kind, "wkt")) info$wkt             <- ad$value
    if (identical(ad$kind, "pra")) info$programarea_key <- ad$value
    info
  })
)))
```

:::

## Map

Areas of interest colored by mean sensitivity score using a Spectral color ramp (red = high, blue = low). Scores are ecoregionally rescaled to a 0--100 range within each BOEM Ecoregion, so they reflect relative sensitivity within a region rather than absolute values across regions.

```{r}
#| label: map-html
#| eval: !expr is_html
# mapgl::maplibre() is used instead of mapboxgl(), which does not
# render inside Quarto documents (walkerke/mapgl#3).
mapgl::maplibre(
  bounds = sf::st_bbox(all_areas)
) |>
  mapgl::add_fill_layer(
    id           = "areas",
    source       = all_areas,
    fill_color   = mapgl::interpolate(
      column = "score",
      values = seq(score_rng[1], score_rng[2],
                   length.out = length(score_cols)),
      stops  = score_cols),
    fill_opacity = 0.6,
    tooltip      = mapgl::concat(
      mapgl::get_column("label"),
      " — mean score: ",
      mapgl::get_column("score"))) |>
  mapgl::add_symbol_layer(
    id              = "area_labels",
    source          = sf::st_point_on_surface(all_areas),
    text_field      = mapgl::get_column("label"),
    text_size       = 14,
    text_color      = "#000000",
    text_halo_color = "#ffffff",
    text_halo_width = 2) |>
  mapgl::add_legend(
    "Mean score",
    values   = round(score_rng, 1),
    colors   = score_cols,
    position = "bottom-right") |>
  mapgl::add_fullscreen_control(position = "top-left")
```

```{r}
#| label: map-static
#| eval: !expr '!is_html'
#| fig-cap: "Areas of interest, colored by mean sensitivity score."
bb <- sf::st_bbox(all_areas)
dx <- (bb["xmax"] - bb["xmin"]) * 0.1
dy <- (bb["ymax"] - bb["ymin"]) * 0.1
centroids <- suppressWarnings(sf::st_point_on_surface(all_areas))
world     <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
us_states <- rnaturalearth::ne_states(
  country = "united states of america", returnclass = "sf")
ggplot2::ggplot() +
  ggplot2::geom_sf(
    data      = world,
    fill      = "#e8e4dc",
    color     = "#b9b3a6",
    linewidth = 0.2) +
  ggplot2::geom_sf(
    data      = us_states,
    fill      = NA,
    color     = "#9e9689",
    linewidth = 0.15) +
  ggplot2::geom_sf(
    data      = all_areas,
    ggplot2::aes(fill = score),
    alpha     = 0.75,
    color     = "#333333",
    linewidth = 0.4) +
  ggplot2::geom_sf_text(
    data     = centroids,
    ggplot2::aes(label = label),
    size     = 3,
    fontface = "bold") +
  ggplot2::scale_fill_gradientn(
    colors = score_cols,
    limits = score_rng,
    name   = "Mean score") +
  ggplot2::coord_sf(
    xlim   = c(bb["xmin"] - dx, bb["xmax"] + dx),
    ylim   = c(bb["ymin"] - dy, bb["ymax"] + dy),
    expand = FALSE) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid.major = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    legend.position  = "bottom") +
  ggplot2::labs(x = NULL, y = NULL)
```

```{r}
#| label: scores-header
#| output: asis
cat("## Plot of Scores\n\n")
cat("Flower plots where each petal represents a species category ",
    "(bird, coral, fish, invertebrate, mammal, reptile, other, ",
    "primary productivity). Petal length reflects the component ",
    "sensitivity score (0--100) and petal width reflects the component ",
    "weight. The center value is the weighted mean across all ",
    "categories.\n\n", sep = "")
if (is_html) cat("::: {.panel-tabset}\n\n")
res <- purrr::map_chr(areas_data, function(ad) {
  knitr::knit_child(
    "report_area_child.qmd",
    envir  = rlang::env(ad = ad, section = "scores", is_html = is_html,
                        emit_table = emit_table),
    quiet  = TRUE)
})
cat(res, sep = "\n")
if (is_html) cat("\n:::\n")
```

## Table of Scores

Mean component and overall sensitivity scores per area of interest, with the count of raster cells (N cells) included in the analysis. Component scores are ecoregionally rescaled (0--100) averages across all cells in each area.

```{r}
#| label: scores-table
#| results: asis
# rows = areas, columns = component scores (wide-pivoted from each
# ad$scores tibble) plus N cells and the overall mean score.
scores_tbl <- purrr::map_dfr(areas_data, function(ad) {
  wide <- ad$scores |>
    dplyr::select(component, score) |>
    tidyr::pivot_wider(names_from = component, values_from = score)
  tibble::tibble(
    Area      = ad$label,
    `N cells` = nrow(ad$cells),
    !!!wide,
    Overall   = msens::mean_score(ad$scores))
}) |>
  dplyr::mutate(dplyr::across(
    where(is.numeric) & !`N cells`,
    \(x) round(x, 0))) |>
  dplyr::mutate(`N cells` = scales::comma(`N cells`))

emit_table(
  scores_tbl,
  caption = "Mean component and overall scores per area.",
  align   = c("l", "r", rep("r", ncol(scores_tbl) - 2)))
```

```{r}
#| label: species-header
#| output: asis
cat("\n## Summary of Species\n\n")
cat("Species counts by category and extinction risk, the top 20 ",
    "highest-scoring species (ranked by habitat-weighted extinction ",
    "risk), and a link to download the full species list as CSV. ",
    "Extinction risk categories are consolidated: FWS and NMFS codes ",
    "combined as USA; parenthetical numbers indicate the extinction ",
    "risk score (1\u2013100) used in sensitivity weighting. ",
    "\"other\" includes IUCN:DD, IUCN:LC, and species with no ",
    "assigned risk code (NA = not available).\n\n", sep = "")
if (is_html) cat("::: {.panel-tabset}\n\n")
res <- purrr::map_chr(areas_data, function(ad) {
  knitr::knit_child(
    "report_area_child.qmd",
    envir  = rlang::env(
      ad         = ad,
      section    = "species",
      is_html    = is_html,
      emit_table = emit_table,
      api_base   = api_base,
      mapsp_base = mapsp_base,
      ver        = params$ver),
    quiet  = TRUE)
})
cat(res, sep = "\n")
if (is_html) cat("\n:::\n")
```

```{r}
#| include: false
DBI::dbDisconnect(con, shutdown = TRUE)
```

::: {.callout-caution collapse="true"}
## Software

```{r}
#| label: session-info
#| comment: ""
devtools::session_info()
```

:::
